Cross-linguistic complexity analysis using UD treebanks

### 2023-03-31 paper/paper.tex
update abstract
add email
finalize paper

\author{Aron Winkler \\
  University of Tübungen \\
  \texttt{email@domain}}

\begin{document}
\maketitle
\begin{abstract}
    The aim of this project is to explore the use of UD treebanks in cross-linguistic complexity research. While it is not this paper's objective to make strong theoretical claims on complexity in its own right, it examines complexity measures computed on various UD treebanks across 4 languages. This work focuses on treebanks for English, German, Hungarian, and Chinese, in an attempt to determine whether complexity measures stay consistent within treebanks for the same language, and whether some languages consistently show higher complexity than others.
    These claims are verified with official UD corpora, as well as with a parallel corpus of 50 sentences for each language mentioned above.
\end{abstract}

@@

What constitutes linguistic complexity is an elusive multi-faceted inquiry, and quantifying it is equally challenging. Previous approaches have included text-based metrics as well as eye-tracking studies (e.g. \citealp{Lee:2007}), but a general consensus has not been reached, especially when it comes to the notion of "overall complexity" of a language. In the case that an experiment produces numeric values as a proxy for complexity, it is not obvious that results computed on two different languages are comparable.

@@

Table~\ref{tab:webdata} shows the list of UD web treebanks used in this experiment. For the English language, which has many treebanks available to it, the Atis\footnote{\url{https://github.com/UniversalDependencies/UD_English-Atis}} treebank, the gum GUM\footnote{\url{https://github.com/UniversalDependencies/UD_English-GUM}} (\citealp{berzak2016tle}), and the EWT\footnote{\url{https://github.com/UniversalDependencies/UD_English-EWT}} treebank (\citealp{silveira14gold}) were ised. For German, this work employs GSD\footnote{\url{https://github.com/UniversalDependencies/UD_German-GSD}} (\citealp{mcdonald-etal-2013-universal}) and HDT\footnote{\url{https://github.com/UniversalDependencies/UD_German-HDT}} (\citealp{borges-volker-etal-2019-hdt}, \citealp{hennig-kohn-2017-dependency}, \citealp{hdtextra}, \citealp{hdtguide}). For Chinese, GSD\footnote{\url{https://github.com/UniversalDependencies/UD_Chinese-GSD}} and PUD\footnote{\url{https://github.com/UniversalDependencies/UD_Chinese-PUD}} were the final choice. For Hungarian, only the Szeged\footnote{\url{https://github.com/UniversalDependencies/UD_Hungarian-Szeged}} treebank (\citealp{szeged}) was available - therefore in order to have at least two treebanks for each language, the sentences annotated in the Szeged treebank were randomly distributed across two new treebanks (hun\_szeged\_1, hun\_szeged\_2).

@@

A balance between morphological metrics (e.g. number of word forms per lemma) and syntactic metrics (e.g. number of clauses per sentence) was one of the objectives of this selection. A number of primitive lexical complexity metrics (e.g. type token ratio, token count) were also included. In this sense, noun to verb ratio is commonly used in readability research, as a preponderance of verbs signals more complex sentence. A similar case can be made for the number of clauses per sentence, as a sentence with many subordinate clauses will be naturally harder to process than a singular main clause.

@@

The metrics \emph{vps} and \emph{wfpl} were highlighted as affected by treebank size in the section above, therefore conferming them on same-size treebanks, however small, might still be a valuable undertaking. Indeed, Hungarian, a language with complex morphology, takes the lead in \emph{wfpl} and Chinese outranks German and English, which have a more constrained word order.

>>>>

\author{Aron Winkler \\
  University of Tübungen \\
  \texttt{aron.winkler@student.uni-tuebingen.de}}

\begin{document}
\maketitle
\begin{abstract}
    The aim of this project is to explore the use of UD treebanks in cross-linguistic complexity research. While it is not this paper's objective to make strong theoretical claims on complexity in its own right, it examines complexity measures computed on various UD treebanks across 4 languages. This work focuses on treebanks for English, German, Hungarian, and Chinese, in an attempt to determine whether complexity measures stay consistent within treebanks for the same language, and whether some languages consistently show higher complexity than others. These claims are verified with official UD corpora, as well as with a parallel corpus of 50 sentences for each language mentioned above.

    The overarching objective is to promote use of UD material in research, as well as encourage data collection and annotation for Hungarian, an interesting language which does not to date have a good amount of data.
\end{abstract}

@@

What constitutes linguistic complexity is an elusive multi-faceted inquiry, and quantifying it is equally challenging. Previous approaches have included text-based metrics as well as eye-tracking studies (e.g. \citealp{Lee:2007}), but a general consensus has not been reached, especially when it comes to the notion of "overall complexity" of a language. In the case that an experiment produces numeric values as a proxy for complexity, it is not obvious that results computed on two different languages or even treebanks are comparable.

@@

Table~\ref{tab:webdata} shows the list of UD web treebanks used in this experiment. For the English language, which has many treebanks available to it, the Atis\footnote{\url{https://github.com/UniversalDependencies/UD_English-Atis}} treebank, the gum GUM\footnote{\url{https://github.com/UniversalDependencies/UD_English-GUM}} (\citealp{berzak2016tle}), and the EWT\footnote{\url{https://github.com/UniversalDependencies/UD_English-EWT}} treebank (\citealp{silveira14gold}) were used. For German, this work employs GSD\footnote{\url{https://github.com/UniversalDependencies/UD_German-GSD}} (\citealp{mcdonald-etal-2013-universal}) and HDT\footnote{\url{https://github.com/UniversalDependencies/UD_German-HDT}} (\citealp{borges-volker-etal-2019-hdt}, \citealp{hennig-kohn-2017-dependency}, \citealp{hdtextra}, \citealp{hdtguide}). For Chinese, GSD\footnote{\url{https://github.com/UniversalDependencies/UD_Chinese-GSD}} and PUD\footnote{\url{https://github.com/UniversalDependencies/UD_Chinese-PUD}} were the final choice. For Hungarian, only the Szeged\footnote{\url{https://github.com/UniversalDependencies/UD_Hungarian-Szeged}} treebank (\citealp{szeged}) was available - therefore in order to have at least two treebanks for each language, the sentences annotated in the Szeged treebank were randomly distributed across two new treebanks (hun\_szeged\_1, hun\_szeged\_2).

@@

A balance between morphological metrics (e.g. number of word forms per lemma) and syntactic metrics (e.g. number of clauses per sentence) was one of the objectives of this selection. A number of primitive lexical complexity metrics (e.g. type token ratio, token count) were also included. In this sense, noun to verb ratio is commonly used in readability research, as a preponderance of verbs signals more complex sentences. A similar case can be made for the number of clauses per sentence, as a sentence with many subordinate clauses will be naturally harder to process than a singular main clause.

@@

The metrics \emph{vps} and \emph{wfpl} were highlighted as affected by treebank size in the section above, therefore conferming them on same-size treebanks, however small, might still be a valuable undertaking. Indeed, Hungarian, a language with complex morphology, takes the lead in \emph{wfpl} and Chinese \emph{vps} outranks German and English , which have a more constrained word order.
